Thesis: draft

### 2020-12-12 draft.tex
continued writing chapter 2

Networks or graphs are conceptually very simple and flexible objects that are made of a set of nodes/vertices $V$ and a set of edges/links $E$ where the element of $E$ determine connections between elements of $V$ \cite{Costa2018}
\begin{equation}
	E \subseteq \{\{x, y\}| x, y \in V \}.
\end{equation}
If self-loops are not allowed, we need the following condition on the elements of $E$
\begin{equation}
	E \subseteq \{\{x, y\}| x, y \in V \text{and\ } x \neq y \}.
\end{equation}
A graph with no self-loops is a simple graph. The nodes $x$ and $y$ of an edge $\{x, y\}$ are called the endpoints of the edge. It is possible that nodes are not joined by any edge, such nodes are called isolated. If two or more edges have the same pair of endpoints, the graph is called a multi-graph. This thesis is not concerned with multi-graphs, nor are self-loops allowed. Graphs in which the edges have an orientation are called directed graphs. If the edges have weights $w_{ij}$, one speaks of a weighted graph.\\
A graph can be represented in different ways. One of the most common ways is by use of the adjacency matrix.\\

@@

The adjacency matrix contains a lot of information such as the degree of a node. The degree $k_i$ of a node $i$ is the number of edges attached to the node $i$ or, in others words, $k_i$ represents the number of nearest neighbors of the node $i$. The degree can be obtained from the adjacency matrix in the following way \cite{Mata2020}
\begin{equation}
	k_i = \sum_{j=1}^N A_{ij}.
\end{equation}
In the case of a directed graph one can differentiate between the number of incoming edges $k_i^{\text{in}}$ and the number of outgoing edges $k_i^{\text{out}}$ of the node $i$. These are called the in-degree and out-degree respectively and are defined as \cite{Mata2020}
\begin{equation}
	k_i^{\text{in}} = \sum_{j=1}^N A_{ji} \quad \text{and} \quad k_i^{\text{out}} = \sum_{j=1}^N A_{ij}.
\end{equation}
The total degree of node $i$ is then $k_i = k_i^{\text{in}} + k_i^{\text{out}}$. For weighted graphs the degree is easily generalized to the weighted degree $s_i$, often called strength \cite{Ioannis2007}
\begin{equation}
	s_i = \sum_{j=1}^N w_{ij}.
\end{equation}
From the degree of the nodes in the network one can construct the degree distribution. The degree distribution $P(k)$ represent the fraction of nodes with a degree $k$ or, in other words, the degree distribution gives the probability that a randomly selected node has a degree $k$. The average degree can be obtained from the degree distribution \cite{Mata2020}
\begin{equation}
	\left<k\right> = \frac{1}{N} \sum_{i=1}^N k_i = \sum_k k P(k)
\end{equation}
where $N$ is the total number of nodes in the network. The degree distribution also allows us to classify networks. The two most important classes are homogeneous and heterogeneous networks. Homogeneous networks have a bell curved degree distribution, eg. a Poisson distribution. In this case, most of the nodes have a degree closed to the average degree $k$. Heterogeneous networks, on the other hand, have a power-law degree distribution, $P(k) \sim k^{-\gamma}$. % explain differences  more, consequences of power-law + need to include figure?
They are also referred to as scale-free networks, since they do not posses a characteristic length scale or, in this case, the average degree isn't a characteristic scale for the network. % correct statement?
%maybe explain why we care about degree, centrality measure etc

@@

The clustering coefficient is another important measure of network topology. It is a measure for the number of triangles in a network and determines the connectivity in the neighborhood of a node $i$: if a node $i$ has a high clustering coefficient, its neighbors are likely to be directly connected to each other \cite{Li2017}. Here a triangle is defined as a loop of length three, this is a sequence of nodes $x, y, z, x$ such that $\{x, y\}, \{y, z\}$ and $\{z, x\}$ are edges of the network. The clustering coefficient is thus a way to measure the degree to which nodes in a network tend to cluster \cite{Li2017}. The local clustering coefficient of a node $i$ is defined as
\begin{equation}
	C_i = \frac{2n_i}{k_i(k_i-1)}
\end{equation}
where $n_i$ is the number of edges that actually exist between the nodes in the neighborhood of $i$ and $k_i(k_i-1)/2$ is the maximum number of edges that could exist between them (note that this expression is only valid for an undirected graph, since for a directed graph $e_{ij} \neq e_{ji}$ and we have $k_i(k_i-1)$ possible edges between the neighbors of node i). % should give more mathimatical expression for n_i?
The average clustering coefficient is then given as the average of the local clustering coefficients
\begin{equation}
	\bar{C} = \frac{1}{N}\sum_{i = 1}^N C_i
\end{equation}
where $N$ denotes, once again, the number of nodes in the network. The clustering coefficient is, beside a measure for the connectivity in the network, also linked to the robustness, or resilience against random damage,  of the network \cite{Li2017} \cite{Iyer2013}. \\
A measure that is closely related to the clustering coefficient is the transitivity $T$, defined as \cite{F.Costa2007}
\begin{equation}
	T = \frac{3 \times \text{number of triangles in the network}}{\text{number of connected triples of nodes in the network}}.
\end{equation}
A connected triple is defined as a set of three nodes with at least two or three edges between them. The factor three arises from the fact that each triangle contributes to three different connected triples in the network: one centered at each node in the triangle \cite{F.Costa2007}. 
% talk about difference transitivity and clustering + read https://arxiv.org/pdf/cond-mat/0505185.pdf

\cite{Albert2002} % not used yet, explains statistical mechanics in complex networks

@@

\section{Network models}
% different subsections where you talk about different graph models: erdos-renyi, clustered random, wats strogatz, barabasi albert, etc...

>>>>

Networks or graphs are conceptually very simple and flexible objects that are made of a set of nodes/vertices $V$ and a set of edges/links $E$, where the elements of $E$ determine connections between elements of $V$ \cite{Costa2018}
\begin{equation}
	E \subseteq \{\{x, y\}| x, y \in V \}.
\end{equation}
If self-loops are not allowed, we need the following condition on the elements of $E$
\begin{equation}
	E \subseteq \{\{x, y\}| x, y \in V \text{and\ } x \neq y \}.
\end{equation}
A graph with no self-loops is a simple graph. The nodes $x$ and $y$ of an edge $\{x, y\}$ are called the endpoints of the edge. It is possible that nodes are not joined by any edge, such nodes are called isolated. If two or more edges have the same pair of endpoints, the graph is called a multi-graph. This thesis is not concerned with multi-graphs, nor are self-loops allowed. Graphs in which the edges have an orientation are called directed graphs. If the edges have weights $w_{ij}$, one speaks of a weighted graph.\\
A graph can be represented in different ways. One of the most common ways is by use of the adjacency matrix.\\

@@

The adjacency matrix contains a lot of information such as the degree of a node. The degree $k_i$ of a node $i$ is the number of edges attached to the node $i$ or, in others words, $k_i$ represents the number of nearest neighbors of the node $i$. The degree can be obtained from the adjacency matrix in the following way \cite{Mata2020}
\begin{equation}
	k_i = \sum_{j=1}^N A_{ij}.
\end{equation}
In the case of a directed graph one can differentiate between the number of incoming edges $k_i^{\text{in}}$ and the number of outgoing edges $k_i^{\text{out}}$ of the node $i$. These are called the in-degree and out-degree respectively and are defined as \cite{Mata2020}
\begin{equation}
	k_i^{\text{in}} = \sum_{j=1}^N A_{ji} \quad \text{and} \quad k_i^{\text{out}} = \sum_{j=1}^N A_{ij}.
\end{equation}
The total degree of node $i$ is then $k_i = k_i^{\text{in}} + k_i^{\text{out}}$. For weighted graphs the degree is easily generalized to the weighted degree $s_i$, often called strength \cite{Ioannis2007}
\begin{equation}
	s_i = \sum_{j=1}^N w_{ij}.
\end{equation}
From the degree of the nodes in the network one can construct the degree distribution. The degree distribution $P(k)$ represent the fraction of nodes with a degree $k$ or, in other words, the degree distribution gives the probability that a randomly selected node has a degree $k$. The average degree can be obtained from the degree distribution \cite{Mata2020}
\begin{equation}
	\left<k\right> = \frac{1}{N} \sum_{i=1}^N k_i = \sum_k k P(k)
\end{equation}
where $N$ is the total number of nodes in the network. The degree distribution also allows us to classify networks. The two most important classes are homogeneous and heterogeneous networks. Homogeneous networks have a bell curved degree distribution, eg. a Poisson distribution. In this case, most of the nodes have a degree closed to the average degree $k$. Heterogeneous networks, on the other hand, have a power-law degree distribution, $P(k) \sim k^{-\gamma}$. % explain differences  more, consequences of power-law + need to include figure?
They are also referred to as scale-free networks, since they do not posses a characteristic length scale or, in this case, the average degree isn't a characteristic scale for the network \cite{Mata2020}. % correct statement?
%maybe explain why we care about degree, centrality measure etc

@@

The clustering coefficient is another important measure of network topology. It is a measure for the number of triangles in a network and determines the connectivity in the neighborhood of a node $i$: if a node $i$ has a high clustering coefficient, its neighbors are likely to be directly connected to each other \cite{Li2017}. Here a triangle is defined as a loop of length three, this is a sequence of nodes $x, y, z, x$ such that $\{x, y\}, \{y, z\}$ and $\{z, x\}$ are edges of the network. The clustering coefficient is thus a way to measure the degree to which nodes in a network tend to cluster \cite{Li2017}. The local clustering coefficient of a node $i$ is defined as
\begin{equation}\label{clus}
	C_i = \frac{2n_i}{k_i(k_i-1)}
\end{equation}
where $n_i$ is the number of edges that actually exist between the nodes in the neighborhood of $i$ and $k_i(k_i-1)/2$ is the maximum number of edges that could exist between them (note that this expression is only valid for an undirected graph, since for a directed graph $e_{ij} \neq e_{ji}$ and we have $k_i(k_i-1)$ possible edges between the neighbors of node i). % should give more mathimatical expression for n_i?
The average clustering coefficient is then given as the average of the local clustering coefficients
\begin{equation}\label{avClus}
	\bar{C} = \frac{1}{N}\sum_{i = 1}^N C_i
\end{equation}
where $N$ denotes, once again, the number of nodes in the network. The clustering coefficient is, beside a measure for the connectivity in the network, also linked to the robustness, or resilience against random damage,  of the network \cite{Iyer2013} \cite{Li2017}. \\
A measure that is closely related to the clustering coefficient is the transitivity $T$, defined as (valid for undirected, unweighted networks) \cite{F.Costa2007}
\begin{equation}\label{globalTrans}
	T = \frac{3 \times \text{number of triangles in the network}}{\text{number of connected triples of nodes in the network}}.
\end{equation}
A connected triple is defined as a set of three nodes with at least two edges between them, so that each node can be reached from the other two (either directly or indirectly). The factor three arises from the fact that each triangle contributes to three different connected triples in the network: one centered at each node in the triangle \cite{F.Costa2007}. Let us denote the number of triangles as $N_{\Delta}$ and the number of connected triples as $N_3$. These two numbers can be obtained from the adjacency matrix in the following way \cite{F.Costa2007}
\begin{align}
	N_{\Delta} &= \sum_{k > j > i} a_{ij}a_{ik}a_{jk} \\
	N_3 &= \sum_{k > j > i} (a_{ij}a_{ik} + a_{ji}a_{jk} + a_{ki}a_{kj})
\end{align}
where $a_{ij}$ are the elements of the adjacency matrix. It is possible to define the transitivity of one node as \cite{F.Costa2007}
\begin{equation}\label{trans}
	T_i = \frac{N_{\Delta}(i)}{N_3(i)}
\end{equation}
where $N_{\Delta}(i)$ represents the number of triangles that involve node $i$ and $N_3(i)$ is the number of connected triples with $i$ as the central node \cite{F.Costa2007}
\begin{align}
	N_{\Delta}(i) &= \sum_{k > j} a_{ij}a_{ik}a_{jk} \\
	N_3(i) &= \sum_{k > j} a_{ij}a_{ik}
\end{align}
It is not too hard to see that $N_{\Delta}(i)$ counts the number of edges between the neighbors of $i$ and that $N_3(i)$ is equal to $k_i(k_i-1)/2$, where $k_i$ is the degree of node $i$ and thus represents the number of neighbors of $i$. It is thus obvious that the local clustering coefficient $C_i$ (Eq. (\ref{clus})) of a node and the transitivity $T_i$ (Eq. (\ref{trans})) of a node define the same quantity \cite{F.Costa2007}. In the case of the average clustering coefficient $\bar{C}$ and the global transitivity $T$ this is however not the case. The difference between the two definitions is that the average of Equation (\ref{globalTrans}) gives the same weight to each triangle in the network, whereas Equation (\ref{avClus}) gives the same weight to each node. This may lead to slightly different values since nodes with a higher degree may possibly be involved in a higher number of triangles than nodes with a lower degree \cite{F.Costa2007}.

@@

\section{Network models}

In order to study the topological structures of real-world networks, several theoretical network models have been developed. These theoretical models usually have a simpler representation and well known properties that can be derived mathematically. They are widely studied and used extensively. The study of complex networks relies heavily on the knowledge and understanding of these models. In this section we will review some of the most important ones.

\subsection{The Erd\H{o}s-R\'{e}nyi network}

In 1959 the two mathematicians Paul Erd\H{o}s and Alfr\'{e}d R\'{e}nyi proposed a model to generate random networks \cite{F.Costa2007}. Idependently of them, Solomonoff and Rapoport already proposed the model in 1951. The model constructs random networks with $N$ nodes in the following way \cite{Albert2002}
\begin{enumerate}
	\item Start with $N$ disconnected nodes,
	\item For each pair of nodes, connect them with a predefined probability $p$.
\end{enumerate} 
If $p=1$, we obtain a network with the maximum number of edges $N(N-1)/2$. This is called the fully connected network.\\
The probability that a node is connected to $k$ other nodes and not to $N-k$ others is given by
\begin{equation}
	p^k (1-p)^{N-1-k}.
\end{equation}
The number of ways to choose the $k$ nodes among the $N-1$ possible candidates is given by the binomial coefficient
\begin{equation}
	\binom{N-1}{k} = \frac{(N-1)!}{k!(N-1-k)!}
\end{equation}
and thus the probability of being connected to $k$ other nodes becomes
\begin{equation}
	p_k = \binom{N-1}{k}p^k (1-p)^{N-1-k}.
\end{equation}
% continue expression for large N

% different subsections where you talk about different graph models: erdos-renyi, clustered random, wats strogatz, barabasi albert, etc...
